---
title: "Advanced configuration"
author: "Jan Salecker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Variables and constants definitions

Correctly defining variables within the experiment class object is crucial for creating simdesigns.
The implemented simdesigns have different requirements for variable definitions:

   Simdesign       | Variable requirements              |  data type 
------------------ | ---------------------------------- | -----------
simdesign_simple   | only constants are used            | any
simdesign_distinct | values (need to have equal length) | any
simdesign_ff       | values, or min, max, step (values is prioritized) | any
simdesign_lhs      | min, max, qfun                     | numeric
simdesign_sobol    | min, max, qfun                     | numeric
simdesign_sobol2007 | min, max, qfun                     | numeric
simdesign_soboljansen | min, max, qfun                     | numeric
simdesign_morris   | min, max, qfun                     | numeric
simdesign_eFast    | min, max, qfun                     | numeric
simdesign_genSA    | min, max                     | numeric
simdesign_genAlg    | min, max                     | numeric


Additionally, please note the following restrictions in order to define variables and constants correctly:

* Categorical variable values are currently only allowed for simdesign_simple, simdesign_distinct and simdesign_ff.
* Variable values that should be recognized by NetLogo as strings need to be nested inside escaped quotes (e.g. `"\\"string\\""`).
* Variable values that should be recognized by NetLogo as logical need to be entered as strings (e.g. `"false"`).
* It is not allowed to list the same variable in the variables and constants list. 
* NetLogo model parameters that are not listed in any of these two lists will be set with their default value from the NetLogo model interface.

A complete list of all valid NetLogo parameters can be loaded by commiting a nl object with a valid modelpath to the function `report_model_parameters()`. This function reads all GUI elements of the NetLogo model that can be set by nlrx.
After attaching an experiment to an nl object, validity of defined experiment variables and constants can be checked by commiting the nl object to the function `eval_variables_constants()`. The function will report detailed warnings or error messages, if definitions of variables or constants are invalid.

## Capturing progress of model simulations

The `run_nl_all()`, `run_nl_one()` and `run_nl_dyn()` functions provide a "silent" parameter that allows to capture progress of running simulations.
If silent is set to FALSE, a message with the current random seed and siminputrow is printed to the console after successful execution of each simulation run. In addition, NetLogo print commands are redirected to the R console. Thus, print commands can be used within the NetLogo model code to display the current progress of simulations in the R console. Another possibility is, to define a print reporter procedure in the experiment slot "idfinal" that is executed at the end of each simulation.

However, both approaches only work for sequential execution. Capturing output from multiple processes in parallelized environments to one R console is not straightforward. If such functionality is needed, we suggest to write the current progress to an output file directly from NetLogo (for example using the idrunnum functionality of nlrx, see section "Notes on self-written output"). These output files can then be monitored to capture the progress of the parallelized model executions. 

## Handling NetLogo runtime errors

Usually, runtime errors of NetLogo model simulations are printed to the R console and the current execution of `run_nl_all()`, `run_nl_one()` or `run_nl_dyn()` breaks. However, it can happen that a model simulation freezes due to Java runtime errors. Unfortunately it is not possible to terminate the Java virtual machine or print an error message to the console after such a runtime error occurred.
The current R session and the freezed Java Virtual Machine need to be terminated manually.
Thus, NetLogo models should be debugged in NetLogo prior to execution of large experiments with nlrx.
Capturing progress of model simulations (see section "Capturing progress of model simulations") might help in debugging runtime errors that freeze the Java Virtual Machine.


## Self-written NetLogo output

The experiment provides a slot called "idrunnum".
This slot can be used to transfer the current nlrx experiment name, random seed and runnumber (siminputrow) to NetLogo.
To use this functionality, a string input field widget needs to be created on the GUI of your NetLogo model.
The name of this widget can be entered into the "idrunnum" field of the experiment.
During simulations, the value of this widget is automatically updated with a generated string that contains the current nlrx experiment name, random seed and siminputrow ("expname_seed_siminputrow").
For self-written output In NetLogo, we suggest to include this global variable which allows referencing the self-written output files to the collected output of the nlrx simulations in R.

## Temporary files management

nlrx uses temporary files to store experiment xml files, commandline batch files to start NetLogo simulations and csv output files.
These temporary files are stored in the default temporary files directory of the R session.
By default, these files are deleted after each simulation run. However, if it is needed to look at this files, automatic deletion of temporary files can be disabled by setting the corresponding cleanup parameters in the `run_nl` functions (cleanup.csv, cleanup.xml, cleanup.bat function parameters).

On unix systems, it can happen that system processes delete files in the default temporary files folder. Thus, we recommend to reassign the temporary files folder for the R-session to another folder. The R-package [unixtools](https://www.rforge.net/unixtools/) provides a function to reassign the temporary files folder for the current R-session:
```{r eval=FALSE}
install.packages('unixtools', repos = 'http://www.rforge.net/')
unixtools::set.tempdir("<path-to-temp-dir>")
```


## random-seed and repetitions management

The experiment provides a slot called "repetition" which allows to run multiple simulations of one parameterization.
This is only useful if you manually generate a new random-seed during the setup of your model.
By default, the NetLogo random-seed is set by the simdesign that is attached to your nl object.
If your model does not reset the random seed manually, the seed will always be the same for each repetition.

However, the concept of nlrx is based on sensitivity analyses. Here, you may want to exclude stochasticity from your output and instead do multiple sensitivity analyses with the same parameter matrix but different random seeds. You can then observe the effect of stochasticity on the level of your final output, the sensitivity indices. Thus we suggest to set the experiment repetition to 1 and instead use the nseeds variable of the desired simdesign to run multiple simulations with different random seeds.

In summary, if you set the random-seed of your NetLogo model manually, you can increase the repitition of the experiment to run several simulations with equal parameterization and different random seeds.
Otherwise, set the experiment repetition to 1 and increase the nseeds variable of your desired simdesign.

## Runtime and measurements

The runtime of NetLogo model simulations can be controlled with two slots of the experiment class:

* runtime - an integer that defines the maximum number of ticks that are simulated
* stopcond - a NetLogo reporter that reports true/false. Simulations are automatically stopped if TRUE is reported. Defaults to NA_character_ which means, no stop condition is applied.

Runtime can be set to 0 or NA_integer_ to allow for simulations without a pre-defined maximum runtime.
However, this should only be done in combination with a proper stop condition (stopcond) or with NetLogo models that have a built-in stop condition. Otherwise, simulations might get stuck in endless loops.

Two slots of the experiment class further define when measurements are taken:

* tickmetrics - defines if measurements are taken at the end of the simulation (final tick) or on each tick
* evalticks - only applied when tickmetrics = "true"; defines a vector of ticks for which output metrics are reported. Set to NA_integer_ to report all collected output.

Depending on the evalticks definition, it might happen, that a simulation stops before any output has been collected.
In such cases, output is still reported but all metrics that could not be collected for any defined evalticks will be filled up with NA.

Four slots of the experiment class further define which measurements are taken:

* metrics - vector of valid NetLogo reporters that are used to collect data (e.g. c("count turtles"))
* metrics.turtles - a list with named vectors of strings defining valid turtles-own variables that are taken as output measurements (e.g. list("turtles" = c("who", "pxcor", "pycor", "color"))
* metrics.patches - vector of valid patches-own variables that are used to collect patches data (e.g. c("pxcor", "pycor", "pcolor"))
* metrics.links - a list with named vectors of strings defining valid links-own variables that are taken as output measurements (e.g. list("links" = c("[who] of end1", "[who] of end2")))

Although the metrics slot accepts any valid NetLogo reporter, such as "count patches", reporter strings can become quite long and confusing. We suggest to create NetLogo reporter procedures for complex reporters in order to get a nice and clean results data frame.
For example, the NetLogo reporter "count patches with [pcolor = green]" could be written as a NetLogo reporter function:
```{r eval=FALSE}
to-report green.patches
  report count patches with [pcolor = green]
end
```
In your nlrx experiment metrics field you can then enter "green.patches" which is way more intuitive then "count patches with [pcolor = green]".

## NetLogo extensions

Usually, all NetLogo extensions that are shipped with NetLogo should also work with nlrx.
However, depending on the system it can happen that NetLogo extensions are not found properly. To solve such problems we advise to put your `.nlogo model` file in the `app/models` subdirectory of the NetLogo installation path.
A special case is the NetLogo r-extension because it needs to be stopped manually in between model runs.
To achieve that, simply put the `r:stop` command in the `idfinal` slot of your experiment: `idfinal = "r:stop"`.


## Parallelisation and the future concept

The run_nl_all function uses the `future_map_dfr()` function from the [furrr package](https://github.com/DavisVaughan/furrr). The simulations are executed in a nested loop where the outer loop iterates over the random seeds of your simdesign, and the inner loop iterates over the rows of the siminput parameter matrix of your simdesign. These loops can be executed in parallel by setting up an appropriate plan from the [future package](https://github.com/HenrikBengtsson/future). See examples below for more details on parallelisation on local machines and remote HPC clusters.

### Parallelisation on local machine

Model simulations can be distributed to each logical processor on the local machine in parallel. The future package provides two options for parallelization, explicit futures and implicit futures which are executed in the background and do not block the console while running. 

Running parallel simulations with an explicit future command:
```{r eval=FALSE}
library(future)
plan(multiprocess)
results <- run_nl_all(nl = nl)
```

For running parallel simulations with an implicit future command we need to define the type of parallelisation for each level of the nested `furrr::future_map_dfr()` function individually. Because we want to parallelize the actual simulations, we need to use the multiprocess plan on the inner level.
Note, that we use the assignment operator for implicit futures (`%<-%`):
```{r eval=FALSE}
library(future)
plan(list(sequential, multiprocess))
results %<-% run_nl_all(nl = nl)
```

In cases, where the number of random seeds is lower than the available processor cores, parallelisation may not be completely efficient. To allow efficient parallelisation, even for a small number of random seeds the split parameter of the `run_nl_all()` function can be used to split the parameter matrix into smaller chunks, which can be distributed to separate processor cores. For example, a simulation with 1000 runs (rows of the siminput matrix) and 2 random seeds should be distributed to 8 processor cores. By default, the parallelisation loop would consist of two jobs (one for each random seed) with 1000 simulation runs each. This experiment would only utilize 2 of the 8 available processor cores. By setting the split parameter to 4, we increase the total number of jobs from 2 to 8 (2 random-seeds * 4 parameter matrix chunks). Each job runs 1/4th of the parameter input matrix (250 rows) using one of the 2 defined random seeds.
```{r eval=FALSE}
library(future)
plan(multisession)
results <- run_nl_all(nl = nl, split = 4)
```



### Parallelisation on remote HPC cluster

This option requires access to a remote HPC cluster.
This example gives you some guidance and examples for sending jobs from an R session on your local machine to an HPC running slurm. Details might be different depending on the HPC setup.

Some settings, such as ssh access and slurm templates need to be defined to access remote HPC clusters from your local R session. Please check out this [detailed HPC setup manual](https://r-spatialecology.github.io/gwdg_hpc_guide/) for examples on required settings for an HPC running slurm.

In order to run NetLogo models on remote HPCs, required software needs to be installed on the remote system as well: java, Netlogo, R, nlrx and further required R-packages (future/clusterMQ, ...). Of course the NetLogo model files need to be available on the remote machine as well.

#### Using the future framework

For sending jobs to the remote HPC under the future framework, we need to install and load additional R packages and adjust the future plan before executing `run_nl_all()`.
You need to define a path to your ssh key, a server adress for the HPC, a user name for the HPC and a path to the slurm template file on the HPC.
Please also make sure that the `nlpath`, `modelpath` and `outpath` variables within your nl object point to locations on the filesystem of the HPC and not your local filesystem.

```{r eval=FALSE}
# Load required packages
library(future)
library(future.batchtools)
library(debugme)
Sys.setenv(DEBUGME='batchtools')
library(batchtools)

# Define the path to the ssh key for your machine:
options(future.makeNodePSOCK.rshopts = c("-i", "/patch/to/id_rsa"))
# Define server and your login credentials for the remote HPC:
login <- tweak(remote, workers="server.HPC.de", user="username")

# Define plan for future environment:
bsub <- tweak(batchtools_slurm, template = "slurm.tmpl", # define name of slurm tmeplate on HPC filesystem
              resources = list(job.name = "jobname", # define jobname
                               log.file = "jobname.log", # define logfile name
                               queue = "medium",  # define HPC queue
                               service = "normal", # define HPC service
                               walltime = "00:30", # define walltime
                               n_jobs = "1",   # define number of processes per job
                               mem_cpu = "4000") # define memory per cpu   

# Load HPC plan:
plan(list(login,
          bsub,
          multisession))

# Execute simulations
results <- run_nl_all(nl = nl)
```

#### Using the clusterMQ framework

The clusterMQ framework is somewhat different from the future framework. However, in our experience it worked more reliable in combination with a slurm HPC.
For installation of clustermq and .Rprofile settings see also the [detailed HPC setup manual](https://r-spatialecology.github.io/gwdg_hpc_guide/).

clusterMQ does not directly support parallelisation of the nested `furrr::future_map_dfr()` loops of the `run_nl_all()` function. We need to define our own parallel simulation function, using the `run_nl_one()` function of the nlrx package:

```{r eval=FALSE}
library(clustermq)

# First, we set the total number of jobs for the HPC
# In this example we run each simulation as an individual job (recommended).
# Thus to calculate the number of jobs we just multiply the number of parameterizations of the simdesign with the number of random seeds.
# If you want to group several runs into the same job you can adjust this line and choose a lower number.
# However, the number must be chosen that nrow(nl@simdesign@siminput)/njobs results in an integer value
njobs <- nrow(nl@simdesign@siminput) * length(nl@simdesign@simseeds)

# Second, we generate vectors for looping trough model runs.
# We generate a vector for simpinputrows by repeating the sequence of parameterisations for each seed.
# Then, we generate a vector of random-seeds by repeating each seed for n times, where n is the number of siminputrows.
siminputrows <- rep(seq(1:nrow(nl@simdesign@siminput)), length(nl@simdesign@simseeds))
rndseeds <- rep(nl@simdesign@simseeds, each=nrow(nl@simdesign@siminput))

# Third, we define our simulation function
# Please adjust the path to the temporary file directory
simfun <- function(nl, siminputrow, rndseed, writeRDS=FALSE)
{
  unixtools::set.tempdir("/hpath/to/temp/dir")
  library(nlrx)
  res <- run_nl_one(nl = nl, siminputrow = siminputrow, seed = rndseed)
  if (isTRUE(writeRDS))
  {
    saveRDS(res, file=file.path(nl@experiment@outpath, paste0("nlrx_sim_", siminputrow, "_seed_", rndseed, ".rds")))
  }
  return(res)
}

# Fourth, use the Q function from the clustermq package to run the jobs on the HPC:
# The q function loops through our siminputrows and rndseeds vectors.
# The function creates njobs jobs and distributes corresponding chunks of the input vectors to these jobs for executing the simulation function simfun.
# As constants we provide our nl object and the function parameter writeRDS. 
# If write RDS is true, an *.rds file will be written on the HPC after each jobs has finished.
# This can be useful to gain results from completed runs after a crash has occured.
results <- clustermq::Q(fun = simfun, 
                        siminputrow = siminputrows,
                        rndseed = rndseeds,
                        const = list(nl = nl,
                                     writeRDS = TRUE),
                        export = list(), 
                        seed = 42, 
                        n_jobs = njobs, 
                        template = list(job_name = "jobname", # define jobname
                                        log.file = "jobname.log", # define logfile name
                                        queue = "medium",  # define HPC queue
                                        service = "normal", # define HPC service
                                        walltime = "00:30:00", # define walltime
                                        mem_cpu = "4000")) # define memory per cpu   

# The Q function reports the individual results of each job as a list
# Thus, we convert the list results to tibble format:
results <- dplyr::bind_rows(results)
```
