---
title: "Advanced configuration"
author: "Jan Salecker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Variables and constants definitions

Correctly defining variables within the experiment class object is crucial for creating simdesigns.
The implemented simdesigns have different requirements for variable definitions:

   Simdesign       | Variable requirements              |  data type 
------------------ | ---------------------------------- | -----------
simdesign_simple   | only constants are used            | any
simdesign_distinct | values (need to have equal length) | any
simdesign_ff       | values, or min, max, step (values is prioritized) | any
simdesign_lhs      | min, max, qfun                     | numeric
simdesign_sobol    | min, max, qfun                     | numeric
simdesign_sobol2007 | min, max, qfun                     | numeric
simdesign_soboljansen | min, max, qfun                     | numeric
simdesign_morris   | min, max, qfun                     | numeric
simdesign_eFast    | min, max, qfun                     | numeric
simdesign_genSA    | min, max                          | numeric
simdesign_genAlg    | min, max                         | numeric
simdesign_ABCmcmc_Marjoram | min, max, qfun            | numeric
simdesign_ABCmcmc_Marjoram_original | min, max, qfun   | numeric
simdesign_ABCmcmc_Wegmann | min, max, qfun             | numeric


Additionally, please note the following restrictions in order to define variables and constants correctly:

* Categorical variable values are currently only allowed for simdesign_simple, simdesign_distinct and simdesign_ff.
* Variable values that should be recognized by NetLogo as strings need to be nested inside escaped quotes (e.g. `"\"string\""`).
* Variable values that should be recognized by NetLogo as logical need to be entered as strings (e.g. `"false"`).
* It is not allowed to list the same variable in the variables and constants list. 
* NetLogo model parameters that are not listed in any of these two lists will be set with their default value from the NetLogo model interface.
* simdesign_simple() requires at least one defined constant within the constants list. If your model does not have any globals (GUI and code), please create a dummy global (either create a global widget on the GUI or add a dummy variable to the globals[] code section) for your model and put it in the constants list with an appropriate value.

A complete list of all valid NetLogo parameters can be loaded by committing a nl object with a valid modelpath to the function `report_model_parameters()`. This function reads all GUI elements of the NetLogo model that can be set by nlrx.
After attaching an experiment to an nl object, validity of defined experiment variables and constants can be checked by commiting the nl object to the function `eval_variables_constants()`. The function will report detailed warnings or error messages, if definitions of variables or constants are invalid.

## Print functions

`print(nl)` can be used with any nl class object. The function will print a formatted overview of class contents to the console.
The function will also print a short summary checklist that may be helpful for debugging certain issues.
Depending on the simdesign, the summary table also prints the estimated number of runs.


## Capturing progress of model simulations

The `run_nl_all()` function supports the [progressr](https://github.com/futureverse/progressr) framework for capturing progress of simulations. Following this logic, the function itself will be silent unless it is wrapped within a `progressr::with_progress()` call. By using different handlers, the layout of the progress bar is different. For example, installing the package [progress](https://github.com/r-lib/progress) and using the handler `progressr::handlers("progress")` will additionally print the current row and seed of the parameter matrix. For example, to run a nl parameter matrix with progress bar one can do:

```{r progress, eval=FALSE}
progressr::handlers("progress")
results <- progressr::with_progress(run_nl_all(nl))
```

The `run_nl_dyn()` function might provide progress output depending on the chosen method (for example ABC offers a progress bar). However, dynamic experiments are difficult to track because it is not clear how long the complete experiment will take from the beginning.

The `run_nl_one()` does not report any progress because it only executes one simulation.

In addition, NetLogo print commands are redirected to the R console. Thus, print commands can be used within the NetLogo model code to display the current progress of simulations in the R console. Another possibility is, to define a print reporter procedure in the experiment slot "idfinal" that is executed at the end of each simulation.

Capturing output from multiple processes in parallelized environments to one R console is not straightforward. If such functionality is needed, we suggest to write the current progress to an output file directly from NetLogo (for example using the idrunnum functionality of nlrx, see section "Notes on self-written output"). These output files can then be monitored to capture the progress of the parallelized model executions. 

## Access to intermediate results

For long-running experiments and simdesigns, it may be beneficial to gain access to intermediate results. 
This is especially useful to obtain results of successful runs, even if the R session crashes or gets stalled.
The `run_all_all()` and `run_nl_one()` functions provide a parameter `writeRDS`. If this is set to `TRUE`, a rds file, carrying the current seed and siminputrow will be written to the specified `outpath` of the nl object experiment. Please note, that in case of the `run_all_all()` function, a huge number of rds files may be written to your disk (check `print(nl)` to get an estimated number of runs/files) which may cause other issues.

Another option to get intermediate results, is to split-up the job into smaller chunks.
E.g. one could create a loop around the `run_nl_all()` and in each iteration only use part of the `nl@simdesign@siminput` tibble (temporary overwrite the `nl@simdesign@siminput` tibble with a copy that only contains a specific slice of the parameter matrix). Then after each iteration, one can write the results of the complete chunk as an rds file to disk.
Another option would be to create an outer loop around the `run_nl_all()` using the random seeds from the `nl@simdesign@simseeds` slot. Then, within each loop iteration the `nl@simdesign@simseeds` slot would be overwritten with the seed from the current loop iteration before the nl object is passed to `run_nl_all()`. Then again, after each iteration, one can write the results of the complete chunk as an rds file to disk.

To collect the intermediate results, simply read all rds files from the outpath in a loop and combine the tibbles, for example by using `dplyr::bind_rows()`.

## Handling NetLogo runtime errors

Usually, runtime errors of NetLogo model simulations are printed to the R console and the current execution of `run_nl_all()`, `run_nl_one()` or `run_nl_dyn()` breaks. However, it can happen that a model simulation freezes due to Java runtime errors. Unfortunately it is not possible to terminate the Java virtual machine or print an error message to the console after such a runtime error occurred.
The current R session and the frozen Java Virtual Machine need to be terminated manually.
Thus, NetLogo models should be debugged in NetLogo prior to execution of large experiments with nlrx.
Capturing progress of model simulations (see section "Capturing progress of model simulations") might help in debugging runtime errors that freeze the Java Virtual Machine.


## Self-written NetLogo output

The experiment provides a slot called "idrunnum".
This slot can be used to transfer the current nlrx experiment name, random seed and runnumber (siminputrow) to NetLogo.
To use this functionality, a string input field widget needs to be created on the GUI of your NetLogo model.
The name of this widget can be entered into the "idrunnum" field of the experiment.
During simulations, the value of this widget is automatically updated with a generated string that contains the current nlrx experiment name, random seed and siminputrow ("expname_seed_siminputrow").
For self-written output In NetLogo, we suggest to include this global variable which allows referencing the self-written output files to the collected output of the nlrx simulations in R.

## Temporary files management

nlrx uses temporary files to store experiment xml files, commandline batch files to start NetLogo simulations and csv output files.
These temporary files are stored in the default temporary files directory of the R session.
By default, these files are deleted after each simulation run. However, if it is needed to look at this files, automatic deletion of temporary files can be disabled by setting the corresponding cleanup parameters in the `run_nl` functions (cleanup.csv, cleanup.xml, cleanup.bat function parameters).

On unix systems, it can happen that system processes delete files in the default temporary files folder. Thus, we recommend to reassign the temporary files folder for the R-session to another folder. The R-package [unixtools](https://www.rforge.net/unixtools/) provides a function to reassign the temporary files folder for the current R-session:
```{r eval=FALSE}
install.packages('unixtools', repos = 'http://www.rforge.net/')
unixtools::set.tempdir("<path-to-temp-dir>")
```


## random-seed and replication management

Running the same parameterization more than once is how stochasticity is quantified in nlrx.
Every simulation is executed with an explicit random seed, so that each run can be identified and reproduced.

The `nseeds` argument of the simdesign helper functions controls this.
nlrx draws `nseeds` random seeds, stores them in the simdesign (`nl@simdesign@simseeds`), and passes each of them to NetLogo through BehaviorSpace's special `random-seed` variable.
Every parameterization is therefore executed once per seed, the seed is reported back in the `random-seed` column of the results, and re-running a design with the same seeds reproduces it exactly.

```{r eval=FALSE}
# Replicate every parameterization with five different random seeds:
nl@simdesign <- simdesign_lhs(nl = nl, samples = 100, nseeds = 5, precision = 3)
```

The concept of nlrx is based on sensitivity analyses.
Here, you may want to exclude stochasticity from your output and instead do multiple sensitivity analyses with the same parameter matrix but different random seeds.
You can then observe the effect of stochasticity on the level of your final output, the sensitivity indices.

The seed vector can also be replaced with your own, for example to repeat a specific set of runs:

```{r eval=FALSE}
setsim(nl, "simseeds") <- c(1234, 5678, 9012)
```

A single parameterization can be replicated directly with `run_nl_one()`, by passing more than one seed.
The replicated runs are executed within one NetLogo instance and can be parallelized with `threads`:

```{r eval=FALSE}
results <- run_nl_one(nl = nl,
                      seed = getsim(nl, "simseeds"),
                      siminputrow = 1,
                      threads = 4)
```

For the dynamic simdesigns (`simdesign_GenSA`, `simdesign_GenAlg` and the `simdesign_ABCmcmc_*` family), replication works differently, because the parameterizations are generated during execution.
There, the `nreplicates` argument of `run_nl_dyn()` defines how often each proposed parameterization is simulated before the results are aggregated and reported back to the algorithm.
See the "Optimization" vignette for details.

> The `repetition` slot of the experiment was removed in nlrx 0.5.0.
> It was passed to BehaviorSpace, which repeated each parameterization with seeds that nlrx neither controlled nor observed, so those runs could not be reproduced.
> Use `nseeds` instead, or `nreplicates` for dynamic designs.

## Runtime and measurements

The runtime of NetLogo model simulations can be controlled with two slots of the experiment class:

* runtime - an integer that defines the maximum number of ticks that are simulated
* stopcond - a NetLogo reporter that reports true/false. Simulations are automatically stopped if TRUE is reported. Defaults to NA_character_ which means, no stop condition is applied.

Runtime can be set to 0 or NA_integer_ to allow for simulations without a pre-defined maximum runtime.
However, this should only be done in combination with a proper stop condition (stopcond) or with NetLogo models that have a built-in stop condition. Otherwise, simulations might get stuck in endless loops.

Two slots of the experiment class further define when measurements are taken:

* tickmetrics - defines if measurements are taken at the end of the simulation (final tick) or on each tick
* evalticks - only applied when tickmetrics = "true"; defines a vector of ticks for which output metrics are reported. Set to NA_integer_ to report all collected output.

Depending on the evalticks definition, it might happen, that a simulation stops before any output has been collected.
In such cases, output is still reported but all metrics that could not be collected for any defined evalticks will be filled up with NA.

Four slots of the experiment class further define which measurements are taken:

* metrics - vector of valid NetLogo reporters that are used to collect data (e.g. c("count turtles"))
* metrics.turtles - a list with named vectors of strings defining valid turtles-own variables that are taken as output measurements (e.g. list("turtles" = c("who", "pxcor", "pycor", "color"))
* metrics.patches - vector of valid patches-own variables that are used to collect patches data (e.g. c("pxcor", "pycor", "pcolor"))
* metrics.links - a list with named vectors of strings defining valid links-own variables that are taken as output measurements (e.g. list("links" = c("[who] of end1", "[who] of end2")))

Although the metrics slot accepts any valid NetLogo reporter, such as "count patches", reporter strings can become quite long and confusing. We suggest to create NetLogo reporter procedures for complex reporters in order to get a nice and clean results data frame.
For example, the NetLogo reporter "count patches with [pcolor = green]" could be written as a NetLogo reporter function:
```{r eval=FALSE}
to-report green.patches
  report count patches with [pcolor = green]
end
```
In your nlrx experiment metrics field you can then enter "green.patches" which is way more intuitive then "count patches with [pcolor = green]".

## NetLogo extensions

Usually, all NetLogo extensions that are shipped with NetLogo should also work with nlrx.
However, depending on the system it can happen that NetLogo extensions are not found properly. To solve such problems we advise to put your `.nlogox` model file in the `models` subdirectory of the NetLogo installation path.
A special case is the NetLogo r-extension because it needs to be stopped manually in between model runs.
To achieve that, simply put the `r:stop` command in the `idfinal` slot of your experiment: `idfinal = "r:stop"`.

## Parallelisation

Since nlrx 0.5.0, parallelisation no longer relies on the `future` and `furrr`
packages. Instead, `run_nl_all()` organises the simulations into **blocks** and
lets NetLogo execute the runs within a block in parallel via its own
multithreading. Two arguments control this:

* `block_size` — how many simulations are bundled into a single NetLogo instance
  (one BehaviorSpace invocation). Blocks are processed one after another, and
  NetLogo is re-initialised once per block.
* `threads` — how many of the runs inside a block NetLogo executes concurrently.
  Check how many are available on your machine with `parallel::detectCores()`.

Note that there is no longer any R-level parallelism *across* blocks: blocks run
sequentially, and all concurrency happens inside NetLogo, controlled by
`threads`.

### Parallelisation on a local machine

To make use of several cores on your machine, bundle many runs into each block
and let NetLogo run them concurrently:

```{r eval=FALSE}
# Use up to 8 concurrent runs, bundling 500 simulations per NetLogo instance:
results <- run_nl_all(nl, block_size = 500, threads = 8)
```

A larger `block_size` means fewer NetLogo restarts (generally faster, with
rapidly diminishing returns) but less frequent progress updates. A smaller
`block_size` updates progress more often but re-initialises NetLogo more
frequently (see also "Capturing progress of model simulations" above).

### Parallelisation on a remote HPC cluster

For large simulation designs, the runs can be distributed across the nodes of an
HPC cluster. The recommended approach uses the
[rslurm](https://cran.r-project.org/package=rslurm) package, which submits an R
function as a SLURM **job array**: each array task runs on its own node and
executes part of the simulation design. Within each node, `block_size` and
`threads` control execution exactly as on a local machine.

The example below splits the parameter table of an `nl` object into a chosen
number of chunks and sends **one chunk per array task**. Each node executes its
chunk with `run_nl_all()` on a sliced copy of the `nl` object, so `block_size`
and `threads` still control execution within the node. This keeps the number of
scheduler jobs modest (one per chunk rather than one per run) — pick `n_chunks`
to match how many nodes you want to occupy.

Prerequisites on the HPC: NetLogo (>= 7), Java, R, and the `nlrx` and `rslurm`
packages must be installed on the cluster, and the `nlpath`, `modelpath` and
`outpath` slots of your `nl` object must point to locations on the HPC
filesystem — not your local machine.

```{r eval=FALSE}
library(rslurm)

# Split the rows of the parameter table into contiguous chunks — one per node:
n_chunks <- 20
row_splits <- split(
  seq_len(nrow(nl@simdesign@siminput)),
  cut(seq_len(nrow(nl@simdesign@siminput)), n_chunks, labels = FALSE)
)

# One row per chunk, storing its first and last siminput row index:
chunks <- data.frame(
  first = vapply(row_splits, min, integer(1)),
  last  = vapply(row_splits, max, integer(1))
)

# The function executed on each node — runs one chunk of the design.
# All random seeds are run for the chunk's parameter rows.
sim_fun <- function(first, last) {
  library(nlrx)
  nl_chunk <- nl
  nl_chunk@simdesign@siminput <- nl@simdesign@siminput[first:last, ]
  run_nl_all(nl_chunk, block_size = 100, threads = 1)
}

sjob <- slurm_apply(
  f              = sim_fun,
  params         = chunks,
  jobname        = "nlrx_sim",
  nodes          = nrow(chunks),        # one array task per chunk
  cpus_per_node  = 1,                   # raise to match `threads` below
  global_objects = "nl",               # ship the nl object to every node
  pkgs           = c("nlrx"),
  slurm_options  = list(
    partition     = "medium",
    time          = "01:00:00",
    "mem-per-cpu" = "2G"
  ),
  submit = FALSE                        # create the job scripts without submitting
)
```

To use NetLogo's multithreading within each node, raise `threads` in `sim_fun()`
and set `cpus_per_node` to the same value.

Setting `submit = FALSE` creates a local `_rslurm_nlrx_sim/` folder with the
generated job scripts without submitting them. Transfer this folder to the HPC,
submit it there with `sbatch`, and once all array tasks have finished, collect
the results (each chunk returns a results tibble, which are row-bound):

```{r eval=FALSE}
results <- get_slurm_out(sjob, outtype = "table")
```

For a complete walk-through of an HPC setup — cluster access, environment,
running R jobs, and NetLogo/nlrx via Apptainer containers — see the
[GWDG HPC guide](https://r-spatialecology.github.io/gwdg_hpc_guide/). The
rslurm-based pattern generalises to any SLURM cluster.
