---
title: "Get Started"
author: "Jan Salecker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

~~You can install the released version of nlrx from~~
~~[CRAN](https://CRAN.R-project.org) with:~~
~~install.packages("nlrx")~~

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nldoc/nlrx")
```

## Example

The nlrx package uses S4 classes to store basic information on NetLogo, the model experiment and the simulation design.
Experiment and simulation design class objects are stored within the nl class object.
This allows to have a complete simulation setup within one single R object.

The following steps guide you trough the process on how to setup, run and analyze NetLogo model simulations with nlrx

#### Step 1: Create a nl object:

The nl object holds all information on the NetLogo version, a path to the NetLogo directory with the defined version, a path to the model file, and the desired memory for the java virtual machine.

```{r eval=FALSE}
nl <- nl(nlversion = "6.0.3",
         nlpath = "C:/Program Files/NetLogo 6.0.3/",
         modelpath = "C:/Program Files/NetLogo 6.0.3/app/models/Sample Models/Biology/Wolf Sheep Predation.nlogo",
         jvmmem = 1024)
```

#### Step 2: Attach an experiment

The experiment object is organized in a similar fashion as NetLogo BehaviorSpace experiments.
It holds information on model variables, constants, metrics, runtime, ...

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep",
                            outpath="C:/out/",
                            repetition=1,
                            tickmetrics="true",
                            idsetup="setup",
                            idgo="go",
                            idfinal=NA_character_,
                            idrunnum=NA_character_,
                            runtime=50,
                            evalticks=seq(40,50),
                            metrics=c("count sheep", "count wolves", "count patches with [pcolor = green]"),
                            variables = list('initial-number-sheep' = list(min=50, max=150, step=10, qfun="qunif"),
                                             'initial-number-wolves' = list(min=50, max=150, step=10, qfun="qunif")),
                            constants = list("model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))
```

#### Step 3: Attach a simulation design

While the experiment defines the variables and specifications of the model, the simulation design creates a parameter input table based on these model specifications and the chosen simulation design method.
nlrx provides a bunch of different simulation designs, such as full-factorial, latin-hypercube, sobol, morris and eFast.
A simulation design is attached to a nl object by using one of these simdesign functions:

```{r eval=FALSE}
nl@simdesign <- simdesign_lhs(nl=nl,
                               samples=100,
                               nseeds=3,
                               precision=3)
```

#### Step 4: Run simulations

All information that is needed to run the simulations is now stored within the nl object.
The run_nl_one() function allows to run one specific simulation from the siminput parameter table.
The run_nl_all() function runs a loop over all simseeds and rows of the parameter input table siminput.
The loops are created by calling furr::future_map_dfr which allows running the function either locally or on remote HPC machines.

```{r eval=FALSE}
future::plan(multisession)

results %<-% run_nl_all(nl = nl, cleanup = "all")
```

#### Step 5: Attach results to nl and run analysis

nlrx provides method specific analysis functions for each simulation design.
Depending on the chosen design, the function reports a tibble with aggregated results or sensitivity indices.
In order to run the analyze_nl function, the simulation output has to be attached to the nl object first.
After attaching the simulation results, these can also be written to the defined outpath of the experiment object.

```{r eval=FALSE}
# Attach results to nl object:
setsim(nl, "simoutput") <- results

# Write output to outpath of experiment within nl
write_simoutput(nl)

# Do further analysis:
analyze_nl(nl)
``` 

## Further notes

#### Comments on simdesigns and variable definitions

Correctly defining variables within the experiment class object is crucial for creating simdesigns.
The implemented simdesigns have different requirements for variable definitions:

   Simdesign       | Variable requirements              |  data type 
------------------ | ---------------------------------- | -----------
simdesign_simple   | only constants are used            | any
simdesign_distinct | values (need to have equal length) | any
simdesign_ff       | values, or min, max, step (values is prioritized) | any
simdesign_lhs      | min, max, qfun                     | numeric
simdesign_sobol    | min, max, qfun                     | numeric
simdesign_sobol2007 | min, max, qfun                     | numeric
simdesign_soboljansen | min, max, qfun                     | numeric
simdesign_morris   | min, max, qfun                     | numeric
simdesign_eFast    | min, max, qfun                     | numeric
simdesign_genSA    | min, max                     | numeric
simdesign_genAlg    | min, max                     | numeric

Categorical variable values are currently only allowed for simdesign_simple, simdesign_distinct and simdesign_ff.
Variable values that should be recognized by NetLogo as strings need to be nested inside escaped quotes (e.g. "\\"string\\"").
Variable values that should be recognized by NetLogo as logical need to be entered as strings (e.g. "false").

#### Comments on self-written output

The experiment provides a slot called "idrunnum".
This slot can be used to transfer the current nlrx experiment name, random seed and runnumber (siminputrow) to NetLogo.
To use this functionality, a string input field widget needs to be created on the GUI of your NetLogo model.
The name of this widget can be entered into the "idrunnum" field of the experiment.
During simulations, the value of this widget is automatically updated with a generated string that contains the current nlrx experiment name, random seed and siminputrow ("expname_seed_siminputrow").
For self-written output In NetLogo, we suggest to include this global vairable which allows referencing the self-written output files to the collected output of the nlrx simulations in R.

#### Comments on seed management

The experiment provides a slot called "repetition" which allows to run multiple simulations of one parameterization.
This is only useful if you manually generate a new random-seed during the setup of your model.
By default, the NetLogo random-seed is set by the simdesign that is attached to your nl object.
If your model does not reset the random seed manually, the seed will always be the same for each repetition.

However, the concept of nlrx is based on sensitivity analyses. Here, you may want to exclude stoachsticity from your output and instead do multiple sensitivity analyses with the same parameter matrix but different random seeds. You can then observe the effect of stochasticity on the level of your final output, the sensitivity indices. Thus we suggest to set the experiment repetition to 1 and instead use the nseeds variable of the desired simdesign to run multiple simulations with different random seeds.

In summary, if you set the random-seed of your NetLogo model manually, you can increase the repitition of the experiment to run several simulations with equal parameterization and different random seeds.
Otherwise, set the experiment repetition to 1 and increase the nseeds variable of your desired simdesign.

#### Comments on measurements

Three slots of the experiment class define how measurements are taken:

* tickmetrics, defines if measurements are taken at the end of the simulation or on each tick

* evalticks, if tickmetrics = "true" evalticks can be used to filter the results for defined ticks

* metrics, definition of valid NetLogo reporters that are used to collect data

Due to the evalticks definition, it might happen, that a simulation stops before any output has been collected.
In such cases, output is still reported but all metrics that could not be collected for any defined evalticks will be filled up with NA.

Although the metrics slot accepts any valid NetLogo reporter, such as "count patches", reporter strings can become quite long and confusing. We suggest to create NetLogo reporter procedures for complex reporters in order to get a nice and clear results data frame.
For example, the NetLogo reporter "count patches with [pcolor = green]" could be written as a NetLogo reporter function:
```{r eval=FALSE}
to-report green.patches
  report count patches with [pcolor = green]
end
```
In your nlrx experiment metrics field you can then enter "green.patches" which is way more intuitive then "count patches with [pcolor = green]".


#### Further notes on the future concept

The run_nl_all function uses the map_dfr functions from the furrr package. The simulations are executed in a nested loop where the outer loop iterates over the random seeds of your simdesign, and the inner loop iterates over the rows of the siminput parameter matrix of your simdesign.

We suggest to always use a future operator (%<-%) when you call this function:
```{r eval=FALSE}
future::plan(multisession)

results %<-% run_nl_all(nl = nl, cleanup = "all")
```
