---
title: "Optimization"
author: "Jan Salecker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Optimization with nlrx

Here we present two simple examples for running an optimization algorithm on a NetLogo model with nlrx.
In our example, we use the Simulated Annealing simdesign (`simdesign_GenSA()`).
However, except for the parameter definitions in the simdesign function and the output of the function, the genetic algorithm optimization (`simdesign_GenAlg()`) works in the same way.

We use the Wolf Sheep Predation model from the models library to show a basic example of the optimization workflow.
Example 1 shows, how a NetLogo reporter can be used as a fitness criterion for optimization. Example 2 uses a self-defined evaluation function that calculates landscape metrics that are then used as fitness criterion.

## Example 1: NetLogo reporter as fitness criterion

#### Step 1: Create a nl object:

```{r eval=FALSE}
library(nlrx)
# Windows default NetLogo installation path (adjust to your needs!):
netlogopath <- file.path("C:/Program Files/NetLogo 7.0.4")
modelpath <- file.path(netlogopath, "models/Sample Models/Biology/Wolf Sheep Predation.nlogox")
outpath <- file.path("C:/out")
# Unix default NetLogo installation path (adjust to your needs!):
netlogopath <- file.path("/home/NetLogo 7.0.4")
modelpath <- file.path(netlogopath, "models/Sample Models/Biology/Wolf Sheep Predation.nlogox")
outpath <- file.path("/home/out")

nl <- nl(nlversion = "7.0.4",
         nlpath = netlogopath,
         modelpath = modelpath,
         jvmmem = 1024)
```

#### Step 2: Attach an experiment

Because we want to apply an optimization algorithm, we need to define proper variable ranges.
The algorithm is allowed to change the values of these parameters within these ranges in order to minimize our fitness criterion.
In this example we want to use a reporter from the metrics slot for evaluating our model runs.
Here we want to find a parameterization that leads to the maximum number of wolfs after 50 ticks.
Because the algorithm automatically searches for minimum values, we add `"1 / count wolves"` to the metrics vector in order to find the maximum number of wolves.

It is also important to think about the settings for tickmetrics, runtime and evalticks.
Because we only want to consider the last tick of the simulation, we set tickmetrics to "false" and runtime to 50. If more than one tick would be measured, the algorithm automatically calculates the mean value of the selected reporter. If you wish to apply other functions to aggregate temporal information into one value, you can use a self-defined evaluation function (see Example 2).

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep-GenSA1",
                            outpath=outpath,
                            tickmetrics="false",
                            idsetup="setup",
                            idgo="go",
                            runtime=50,
                            metrics=c("(1 / count wolves)"),
                            variables = list('initial-number-sheep' = list(min=50, max=150, qfun="qunif"),
                                             'initial-number-wolves' = list(min=50, max=150, qfun="qunif")),
                            constants = list("model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))

```

#### Step 3: Attach a simulation design

We use the `simdesgin_GenSA()` function to attach a Simulated Annealing simdesign.
We select the evaluation criterion (`evalcrit`) by assigning the position of the reporter that we want to evaluate within the metrics vector of the experiment.
In our case, there is only one reporter in the metrics vector thus we set evalcrit to use the first reporter (`evalcrit = 1`).
The control parameter allows us to provide additional parameters for the GenSA function (see ?GenSA for details). For demonstration purposes, we set the maximum number of iterations to 20.

```{r eval=FALSE}
nl@simdesign <- simdesign_GenSA(nl, 
                                evalcrit = 1, 
                                nseeds = 1, 
                                control=list(maxit = 20))
```

#### Step 4: Run simulations

For optimization simdesign, the `run_nl_dyn()` function lets you execute the simulations.
There are some notable differences between `run_nl_all()` and `run_nl_dyn()`.
First, because parameterizations depend of results from previous runs, a single optimization can not be parallelized.
Second, `run_nl_dyn()` runs one complete optimization for each random seed of the simdesign, one after another, and these optimizations are independent of each other.
Third, the results of the optimizations are not returned in one tibble of model output, because optimization procedures do not report model output but their own result objects.
Instead, the returned tibble contains one row per seed, with the seed in the `seed` column and the result object of the respective optimization in the `result` list column.
The objects are stored exactly as the optimization packages return them, so that their summary and plot functions can be applied to them.

```{r eval=FALSE}
results <- run_nl_dyn(nl)
results$result[[1]]
```

If an optimization fails, the loop is not interrupted: the error is stored in the `result` column of that seed, a warning lists the failed seeds, and the results of the remaining seeds are returned.
To run only some of the seeds, for example to distribute them over the nodes of a cluster, reduce the seed vector of the simdesign before the call:

```{r eval=FALSE}
setsim(nl, "simseeds") <- getsim(nl, "simseeds")[1]
```

Optimization algorithms evaluate a stochastic model, so every evaluation of the fitness criterion carries simulation noise.
The `nreplicates` argument runs each proposed parameterization several times with different random seeds and reports the aggregated value to the algorithm.
The values are first averaged over all measured ticks of a run, then over the replicated runs.
Because the replicated runs of one evaluation are executed within a single NetLogo instance, they can be parallelized with `threads`:

```{r eval=FALSE}
results <- run_nl_dyn(nl,
                      nreplicates = 5,
                      threads = 5)
```

The replicate seeds are derived from the seed of the respective optimization and are the same for every evaluation of that optimization.
Reusing the seeds this way means that a difference between two proposed parameterizations reflects the parameters rather than the random draw, which makes the fitness criterion easier to optimize.
It also means that the optimum is found for this particular set of random realisations, so it is good practice to re-evaluate the resulting parameterization with different seeds.

#### Step 5: Investigate output

Each element of the `result` column is the output list of the Simulated Annealing procedure and contains four elements:
`value` reports the minimum final value of the evaluation criterion.
`par` reports the parameter settings of the final parameterisation in the same order as defined in the experiment of the nl object.
`trace.mat` gives you detailed information on the optimization process over all iterations.
`counts` indicates how often the optimization procedure was executed in total.

```{r eval=FALSE}
results
results$result[[1]]$value
results$result[[1]]$par
```

With several seeds, comparing the optima of the individual optimizations shows whether the algorithm converges to the same solution under different stochasticity:

```{r eval=FALSE}
purrr::map_dbl(results$result, "value")
```

In order to store our results together with the nl object we can attach the returned tibble to it:

```{r eval=FALSE}
setsim(nl, "simoutput") <- results
saveRDS(nl, file.path(nl@experiment@outpath, "genSA_1.rds"))

```



## Example 2: Evaluation function as fitness criterion

#### Step 1: Create a nl object:

```{r eval=FALSE}
library(nlrx)
# Windows default NetLogo installation path (adjust to your needs!):
netlogopath <- file.path("C:/Program Files/NetLogo 7.0.4")
modelpath <- file.path(netlogopath, "models/Sample Models/Biology/Wolf Sheep Predation.nlogox")
outpath <- file.path("C:/out")
# Unix default NetLogo installation path (adjust to your needs!):
netlogopath <- file.path("/home/NetLogo 7.0.4")
modelpath <- file.path(netlogopath, "models/Sample Models/Biology/Wolf Sheep Predation.nlogox")
outpath <- file.path("/home/out")

nl <- nl(nlversion = "7.0.4",
         nlpath = netlogopath,
         modelpath = modelpath,
         jvmmem = 1024)
```

#### Step 2: Attach an experiment

Because we want to apply an optimization algorithm, we need to define proper variable ranges.
The algorithm is allowed to change the values of these parameters within these ranges in order to minimize our fitness criterion.
In this example we want to use a self-defined evaluation function to calculate a fitness criterion.
Thus, we add the patch coordinates and patch color (as a patch class indicator) to the `metrics.patches` vector.
We want to use spatial data to calculate the landscape edge density index of the final tick and find a parameterization that leads to the edge density.
Because we only want to consider the last tick of the simulation, we set tickmetrics to "false" and runtime to 50. 

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep-GenSA2",
                            outpath=outpath,
                            tickmetrics="false",
                            idsetup="setup",
                            idgo="go",
                            runtime=50,
                            metrics.patches = c("pxcor", "pycor", "pcolor"),
                            variables = list('initial-number-sheep' = list(min=50, max=150),
                                             'initial-number-wolves' = list(min=50, max=150)),
                            constants = list("model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))

```

#### Step 3: Attach a simulation design

We use the `simdesgin_GenSA()` function to attach a Simulated Annealing simdesign.
Because we want to post-process our simulation results, we need to define an evaluation function.
The evaluation function needs to accept the nl object as input and must return a single numeric value.
First we load the package landscapemetrics. We then convert the spatial data to a raster format and calculate the landscape edge density index.
Finally, we report only the index value of the resulting tibble.

```{r eval=FALSE}
critfun <- function(nl) {
  library(landscapemetrics)
  res_spat <- nl_to_raster(nl)
  res_spat_raster <- res_spat$spatial.raster[[1]]
  lsm <- lsm_l_ed(res_spat_raster)
  crit <- lsm$value
  return(crit)
}
```

In the `simdesign_GenSA()` function we now provide our evaluation function (`critfun`) as evaluation criterion (`evalcrit`).
The control parameter allows us to provide additional parameters for the GenSA function (see ?GenSA for details). For demonstration purposes, we set the maximum number of iterations to 20.

```{r eval=FALSE}
nl@simdesign <- simdesign_GenSA(nl, 
                                evalcrit = critfun, 
                                nseeds = 1, 
                                control=list(maxit = 20))
```

#### Step 4: Run simulations

For optimization simdesign, the `run_nl_dyn()` function lets you execute the simulations.
There are some notable differences between `run_nl_all()` and `run_nl_dyn()`.
First, because parameterizations depend of results from previous runs, a single optimization can not be parallelized.
Second, `run_nl_dyn()` runs one complete optimization for each random seed of the simdesign, one after another, and these optimizations are independent of each other.
Third, the results of the optimizations are not returned in one tibble of model output, because optimization procedures do not report model output but their own result objects.
Instead, the returned tibble contains one row per seed, with the seed in the `seed` column and the result object of the respective optimization in the `result` list column.
The objects are stored exactly as the optimization packages return them, so that their summary and plot functions can be applied to them.

```{r eval=FALSE}
results <- run_nl_dyn(nl)
results$result[[1]]
```

If an optimization fails, the loop is not interrupted: the error is stored in the `result` column of that seed, a warning lists the failed seeds, and the results of the remaining seeds are returned.
To run only some of the seeds, for example to distribute them over the nodes of a cluster, reduce the seed vector of the simdesign before the call:

```{r eval=FALSE}
setsim(nl, "simseeds") <- getsim(nl, "simseeds")[1]
```

Optimization algorithms evaluate a stochastic model, so every evaluation of the fitness criterion carries simulation noise.
The `nreplicates` argument runs each proposed parameterization several times with different random seeds and reports the aggregated value to the algorithm.
The values are first averaged over all measured ticks of a run, then over the replicated runs.
Because the replicated runs of one evaluation are executed within a single NetLogo instance, they can be parallelized with `threads`:

```{r eval=FALSE}
results <- run_nl_dyn(nl,
                      nreplicates = 5,
                      threads = 5)
```

The replicate seeds are derived from the seed of the respective optimization and are the same for every evaluation of that optimization.
Reusing the seeds this way means that a difference between two proposed parameterizations reflects the parameters rather than the random draw, which makes the fitness criterion easier to optimize.
It also means that the optimum is found for this particular set of random realisations, so it is good practice to re-evaluate the resulting parameterization with different seeds.

#### Step 5: Investigate output

The output list of the Simulated Annealing procedure contains four elements:
`value` reports the minimum final value of the evaluation criterion.
`par` reports the parameter settings of the final parameterisation in the same order as defined in the experiment of the nl object.
`trace.mat` gives you detailed information on the optimization process over all iterations.
`counts` indicates how often the optimization procedure was executed in total.

```{r eval=FALSE}
results
results$result[[1]]$value
```

In order to store our results together with the nl object we can attach the returned tibble to it:

```{r eval=FALSE}
setsim(nl, "simoutput") <- results
saveRDS(nl, file.path(nl@experiment@outpath, "genSA_2.rds"))

```






