---
title: "Simdesign examples"
author: "Jan Salecker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Further Wolf Sheep examples for all implemented simdesigns

The following section provides valid experiment setups for all supported simdesigns using the Wolf Sheep Model from the NetLogo models library.

First we set up a nl object. We use this nl object for all simdesigns:

```{r eval=FALSE}
nl <- nl(nlversion = "6.0.3",
         nlpath = "C:/Program Files/NetLogo 6.0.3/",
         modelpath = "C:/Program Files/NetLogo 6.0.3/app/models/Sample Models/Biology/Wolf Sheep Predation.nlogo",
         jvmmem = 1024)
```

##### simdesign_simple

To setup a simple simdesign, no variables have to be defined. The simple simdesign only uses defined constants and reports a parameter matrix with only one parameterization.

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep",
                            outpath="C:/out/",
                            repetition=1,
                            tickmetrics="true",
                            idsetup="setup",
                            idgo="go",
                            idfinal=NA_character_,
                            idrunnum=NA_character_,
                            runtime=50,
                            evalticks=seq(40,50),
                            metrics=c("count sheep", "count wolves", "count patches with [pcolor = green]"),
                            variables = list(),
                            constants = list("initial-number-sheep" = 20,
                                             "initial-number-wolves" = 20,
                                             "model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))

nl@simdesign <- simdesign_simple(nl=nl,
                                 nseeds=3)
```

##### simdesign_distinct

To setup a distinct simdesign, vectors of values need to be defined for each variable. These vectors must have the same number of elemtents across all variables.

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep",
                            outpath="C:/out/",
                            repetition=1,
                            tickmetrics="true",
                            idsetup="setup",
                            idgo="go",
                            idfinal=NA_character_,
                            idrunnum=NA_character_,
                            runtime=50,
                            evalticks=seq(40,50),
                            metrics=c("count sheep", "count wolves", "count patches with [pcolor = green]"),
                            variables = list('initial-number-sheep' = list(values=c(10, 20, 30, 40)),
                                             'initial-number-wolves' = list(values=c(30, 40, 50, 60))),
                            constants = list("model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))

nl@simdesign <- simdesign_distinct(nl=nl,
                                   nseeds=3)
```


##### simdesign_ff

To setup a full-factorial simdesign, vectors of values need to be defined for each variables. Alternatively, a sequence can be defined by setting min, max and step. However, if both (values and min, max, step) are defined, the values vector is prioritized.

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep",
                            outpath="C:/out/",
                            repetition=1,
                            tickmetrics="true",
                            idsetup="setup",
                            idgo="go",
                            idfinal=NA_character_,
                            idrunnum=NA_character_,
                            runtime=50,
                            evalticks=seq(40,50),
                            metrics=c("count sheep", "count wolves", "count patches with [pcolor = green]"),
                            variables = list('initial-number-sheep' = list(values=c(10, 20, 30, 40)),
                                             'initial-number-wolves' = list(min=0, max=50, step=10)),
                            constants = list("model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))

nl@simdesign <- simdesign_ff(nl=nl,
                             nseeds=3)
```

##### simdesign_lhs, _sobol, _sobol2007, _soboljansen, _morris, _eFast

To setup sensitivity analysis simdesigns, variable distributions (min, max, qfun) need to be defined.

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep",
                            outpath="C:/out/",
                            repetition=1,
                            tickmetrics="true",
                            idsetup="setup",
                            idgo="go",
                            idfinal=NA_character_,
                            idrunnum=NA_character_,
                            runtime=50,
                            evalticks=seq(40,50),
                            metrics=c("count sheep", "count wolves", "count patches with [pcolor = green]"),
                            variables = list('initial-number-sheep' = list(min=50, max=150, qfun="qunif"),
                                             'initial-number-wolves' = list(min=50, max=150, qfun="qunif")),
                            constants = list("model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))

nl@simdesign <- simdesign_lhs(nl=nl,
                               samples=100,
                               nseeds=3,
                               precision=3)


nl@simdesign <- simdesign_sobol(nl=nl,
                                 samples=200,
                                 sobolorder=2,
                                 sobolnboot=20,
                                 sobolconf=0.95,
                                 nseeds=3,
                                 precision=3)

nl@simdesign <- simdesign_sobol2007(nl=nl,
                                     samples=200,
                                     sobolnboot=20,
                                     sobolconf=0.95,
                                     nseeds=3,
                                     precision=3)

nl@simdesign <- simdesign_soboljansen(nl=nl,
                                       samples=200,
                                       sobolnboot=20,
                                       sobolconf=0.95,
                                       nseeds=3,
                                       precision=3)


nl@simdesign <- simdesign_morris(nl=nl,
                                  morristype="oat",
                                  morrislevels=4,
                                  morrisr=100,
                                  morrisgridjump=2,
                                  nseeds=3)

nl@simdesign <- simdesign_eFast(nl=nl,
                                 samples=100,
                                 nseeds=3)
```

##### simdesign_GenSA, _GenAlg

To setup optimization simdesigns, variable ranges (min, max) need to be defined.

```{r eval=FALSE}
nl@experiment <- experiment(expname="wolf-sheep",
                            outpath="C:/out/",
                            repetition=1,
                            tickmetrics="true",
                            idsetup="setup",
                            idgo="go",
                            idfinal=NA_character_,
                            idrunnum=NA_character_,
                            runtime=50,
                            evalticks=seq(40,50),
                            metrics=c("count sheep", "count wolves", "count patches with [pcolor = green]"),
                            variables = list('initial-number-sheep' = list(min=50, max=150),
                                             'initial-number-wolves' = list(min=50, max=150)),
                            constants = list("model-version" = "\"sheep-wolves-grass\"",
                                             "grass-regrowth-time" = 30,
                                             "sheep-gain-from-food" = 4,
                                             "wolf-gain-from-food" = 20,
                                             "sheep-reproduce" = 4,
                                             "wolf-reproduce" = 5,
                                             "show-energy?" = "false"))

nl@simdesign <- simdesign_GenAlg(nl=nl, 
                                 popSize = 200, 
                                 iters = 100, 
                                 evalcrit = 1,
                                 elitism = NA, 
                                 mutationChance = NA, 
                                 nseeds = 1)

nl@simdesign <- simdesign_GenSA(nl=nl,
                                par=NULL,
                                evalcrit=1,
                                control=list(max.time = 600),
                                nseeds=1)

```
